---
title: "Clustering RB's"
author: "Jackson Sutkaytis"
date: "2024-11-18"
output: 
  distill::distill_article:
    self_contained: true
    highlight: kate
    toc: true
---

```{r setup, include=FALSE}
# Set options for knitr
knitr::opts_chunk$set(echo = TRUE)

# Loading necessary libraries for data manipulation, visualization, and analysis
library(rmarkdown)
library(ggplot2)
library(nflreadr)
library(tidyverse)
library(gsisdecoder)
library(rvest)
library(janitor)
library(gghighlight)
library(ggrepel)
library(extrafont)
library(gt)
library(ggimage)
library(distill)
library(corrplot)
library(gridExtra)
library(factoextra)
library(ggrepel)
library(dplyr)
```

**Load Rosters, Snap Counts, and Player Stats Data for 2024 Season**
```{r}
#Set Seed
set.seed(2024)

#Each player and their team rosters
rosters <- nflreadr::load_rosters(seasons = 2024)

#Snap counts for each week aggregated for the entire season
seasonSnaps <- nflreadr::load_snap_counts(seasons = 2024)
seasonSnaps <- seasonSnaps[seasonSnaps$position == "RB", ]
seasonSnaps <- seasonSnaps %>% 
  group_by(pfr_player_id, player) %>% 
  summarise(summedSnaps = sum(offense_snaps))

#RB stats like speed and weight
playerStats <- nflreadr::load_combine() %>% filter(pos == "RB")

#All active 32 NFL teams
teams <- nflreadr::load_teams(current = TRUE)
```

**Let's Investigate a Threshold**

```{r warning=FALSE}
#Plotting Snaps by each RB
ggplot(seasonSnaps, aes(x = summedSnaps, y = reorder(player, summedSnaps))) +
  geom_bar(stat = "identity", fill = "orange") +
  labs(
    x = "Player",
    y = "Summed Snaps",
    title = "Summed Snaps by Player"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),  
    plot.title = element_text(size = 14)
  )
#It does not seem like there is a clear fall off, let us just go with the top ~60% eye test. 


#Lets go with 200
JOINTABLE <- seasonSnaps %>% 
  filter(summedSnaps > 200) %>% 
  select(pfr_player_id, player)


```

**Let's figure out what Attributes a RB has that Would be Good for Grouping**

<ol>

1.  40 time

2.  Height

3.  Weight

4.  Yards Per Carry (YPC)

5.  Broken Tackles

6.  Outside or Inside Run Percentage

7.  Avg of YDs after contact

    </ol>

**Now we Will get Some Metrics from Season Data**

```{r warning=FALSE}

#Play Data
DFStats <- load_pbp(seasons = 2024) 

#Wrangle Our Variables to see Run Location
DFStats <- DFStats %>% 
  filter(season_type == "REG" & play_type == "run") %>% 
  select(rusher_player_id, rusher_player_name, rushing_yards, run_location) %>% 
  mutate(run_location = if_else(run_location == "right" | run_location == "left", "outside", run_location)) %>% 
  drop_na()

#Calculate the Metrics
DFStats <- DFStats %>%
  group_by(rusher_player_id, rusher_player_name) %>%
  summarise(
    YPC = mean(rushing_yards),
    total_runs = n(),
    middle_runs = sum(run_location == "middle"),
    middle_run_percent = (middle_runs / total_runs) * 100
  ) %>% 
  select(rusher_player_id, rusher_player_name, YPC, middle_run_percent)

```

**Scraping More Data**

```{r warning=FALSE}
#Random NFL Stats Website
url <- "https://www.fantasypros.com/nfl/advanced-stats-rb.php"
onlineStats <- read_html(url)

#Scrapped the Table
onlineStats <- onlineStats %>% 
  html_node("table") %>% 
  html_table(fill = TRUE)

#Promote Headers
colnames(onlineStats) <- onlineStats[2, ] 
onlineStats <- onlineStats[-c(1, 2), ] 
colnames(onlineStats) <- make.names(colnames(onlineStats), unique = TRUE)

#Touch up the Names and Convert to Keys
onlineStats <- onlineStats %>%
  mutate(Player = gsub("^(\\S+\\s+\\S+)\\s.*", "\\1", Player))

```

**Tie All of our Data Together**

<ol>

1.  DFCombine (pfr_player_id)
2.  DFStats (rusher_player_id)
3.  rosters (gsis_id & pfr_id)
4.  Online Data (player)

</ol>

```{r warning=FALSE}

#Joining Combine Stats
JOINTABLE <- JOINTABLE %>%
    left_join(rosters %>% select(pfr_id, gsis_id), by = c("pfr_player_id" = "pfr_id")) %>%
    left_join(playerStats %>% select(pfr_id, forty, ht, wt), by = c("pfr_player_id" = "pfr_id")) %>%
    rename(
        gsis_id = gsis_id,
        forty = forty,
        ht = ht,
        wt = wt
    ) %>% 
  drop_na()

#Joining Season Stats
JOINTABLE <- JOINTABLE %>% 
  left_join(DFStats %>% select(rusher_player_id, YPC, middle_run_percent), by = c("gsis_id" = "rusher_player_id"))

#Join our Scrapped Data
JOINTABLE <- JOINTABLE %>%
  left_join(
    onlineStats %>% select(Player, BRKTKL, YACON.ATT), 
    by = c("player" = "Player")
  ) %>% 
  drop_na()
#Rename
DATA <- JOINTABLE 

#Convert to Numbers
DATA$BRKTKL <- as.numeric(DATA$BRKTKL)
DATA$YACON.ATT <- as.numeric(DATA$YACON.ATT)


##Convert Height to Numeric
convertToInches <- function(height) {
  parts <- as.numeric(unlist(strsplit(height, "-")))
  return(parts[1] * 12 + parts[2])  
}
DATA$ht <- sapply(DATA$ht, convertToInches)

```

**How About a Little EDA with our Final Table?**
```{r warning=FALSE}

#Look into distributions
DATA %>%
  gather(Attributes, value, c(4:10)) %>% 
  ggplot(aes(x = value, fill = Attributes)) +
  geom_histogram(colour = "black", show.legend = FALSE, bins = 20) + 
  facet_wrap(~Attributes, scales = "free_x") +
  labs(x = "Values", y = "Frequency", title = "Attributes - Histograms") +
  theme_bw()

```

```{r warning=FALSE}
#Density
DATA %>% 
  gather(Attributes, value, c(4:10) ) %>%
  ggplot(aes(x=value, fill=Attributes)) +
  geom_density(colour="black", alpha=0.9, show.legend=FALSE) +
  facet_wrap(~Attributes, scales="free_x") +
  labs(x="Values", y="Density",
       title="Attributes - Density plots") +
  theme_bw()
```

```{r warning=FALSE}
#Subset Numeric Data
numericDATA <- DATA[,c(4:10)]
#Correlation Plot
corrplot(cor(numericDATA, use = "complete.obs"), 
         type = "upper", 
         method = "color", 
         tl.cex = 0.6, 
         addCoef.col = "black") 
```

**Correlations**

```{r warning=FALSE}
#Plotting the Highest Correlation (I Assume Multicolinearity)
ggplot(DATA, aes(x=YPC, y=YACON.ATT)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title="YPC and YACON.ATT") +
  theme_bw()
```

**Scale our data**

```{r}
#Scale and Center
scaledDATA <- prcomp(DATA[, 4:10], center = TRUE, scale. = TRUE)
#Join it Back
DATA[4:10] <- scaledDATA$x

```

**Let the fun begin**

```{r}
#Within Cluster Summed Squares
fviz_nbclust(DATA[, 4:10], kmeans, method = "wss") +
    labs(subtitle = "Elbow Method for K-means Clustering")

##Silhouette Method - Note our value is relatively close to 0, meaning we are having a lot of overlapping clusters. 
fviz_nbclust(DATA[, 4:10], kmeans, method = "silhouette") +
    labs(subtitle = "Silhouette Method for K-means Clustering")


```

**Code for 2 Cores**

```{r}
# Tune
clusterTwo <- kmeans(DATA[, 4:10], centers = 2, nstart = 25)
# Generate the plot
  p2 <- fviz_cluster(clusterTwo, data = DATA[, 4:10],
                    ellipse.type = "euclid", 
                    star.plot = FALSE, 
                    repel = TRUE, 
                    labelsize = 1,
                    ellipse = FALSE,
                    ggtheme = theme_minimal() 
  ) + 
    labs(title = "K-Means 2 Clusters, Euclidian") + 
    theme(plot.title = element_text(size = 10),
          axis.text = element_text(size = 8),  
          axis.title = element_text(size = 10)) 
  
print(p2)

```

**8 Cores**

```{r}
# Tune
clusterEight <- kmeans(DATA[, 4:10], centers = 8, nstart = 25)
# Generate the plot
  p8 <- fviz_cluster(clusterEight, data = DATA[, 4:10],
                    ellipse.type = "euclid", 
                    star.plot = FALSE, 
                    repel = TRUE, 
                    labelsize = 1,
                    ellipse = FALSE,
                    ggtheme = theme_minimal() 
  ) + 
    labs(title = "K-Means 8 Clusters, Euclidian") + 
    theme(plot.title = element_text(size = 10),
          axis.text = element_text(size = 8),  
          axis.title = element_text(size = 10)) 
  
print(p8)
```

**Hierarchical Clustering**

```{r}
#Create the distance matrix
distanceMatrix <- dist(DATA[,4:10], method = "euclidean")

# Perform hierarchical clustering
hclust <- hclust(distanceMatrix, method = "complete")

# Create the dendrogram
dog <- plot(hclust, 
     main = "Hierarchical Clustering Dendrogram", 
     xlab = "Observations", 
     ylab = "Height", 
     col = "blue", 
     cex = 0.6, 
     hang = -1, 
     labels = DATA$player) 

rect.hclust(hclust, k = 8, border = 2:5)


```


**The Grand Plot**
```{r}

#Join Team to Use as a Key
DATA <- DATA %>% 
  left_join(rosters %>%  select(full_name, last_name, team), join_by("player" == "full_name")) 
#Join the Logo URL
DATA <- DATA %>% 
  left_join(teams %>%  select( team_abbr, team_logo_espn), join_by("team" == "team_abbr"))
#Cut our Tree
clusterLabs <- cutree(hclust, k = 8)
#Add our clusters to the DF
DATA$Cluster <- factor(clusterLabs)

#Add Our Principle Component Analysis
DATA$PCA1 <- scaledDATA$x[,1]
DATA$PCA2 <- scaledDATA$x[,2]
#Aggregate our PCAs
cluster_centers <- aggregate(cbind(PCA1, PCA2) ~ Cluster, data = DATA, mean)

#Plot PCAs and Clusters
ggplot(DATA, aes(x = PCA1, y = PCA2)) +
  geom_image(aes(image = team_logo_espn), size = 0.05) +  # Add team logos as points
  geom_text_repel(aes(label = player, color = Cluster), size = 2.5, max.overlaps = 10) +  # Color labels by cluster
  scale_color_manual(values = rainbow(length(unique(DATA$Cluster)))) +  # Distinct colors for clusters
  labs(x = "PCA1 (Most Significant Pattern)", y = "PCA2 (2nd Significant Pattern)", color = "Cluster", title = "Clusters with Team Logos and Player Labels") +
  theme_minimal() +
  theme(legend.position = "none")



```
